VariantSpark: Population Scale Clustering of Genotype Information

### 2015-11-27 doc/publications/journal2014/paper.tex
update to paper and letter

\author[
   addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
%   corref={aff1},                       % id of corresponding address, if any
%   noteref={n1},                        % id's of article notes, if any
%   email={jane.e.doe@cambridge.co.uk}   % email address
]{\inits{}\fnm{Aidan R.} \snm{O'Brien}}
\author[
   addressref={aff1},
]{\inits{}\fnm{Neil F.W.} \snm{Saunders}}
\author[
   addressref={aff3,aff4},
%   email={john.RS.Smith@cambridge.co.uk}
]{\inits{}\fnm{Fabian A.} \snm{Buske}}
\author[
   addressref={aff2},
   email={rodney.scott@newcastle.edu.au}
]{\inits{}\fnm{Rodney J.} \snm{Scott}}
\author[
   addressref={aff1},
   email={Denis.Bauer@CSIRO.au}
%   email={john.RS.Smith@cambridge.co.uk}
]{\inits{}\fnm{Denis C.} \snm{Bauer}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%                           % unique id
  \orgname{CSIRO, Health \& Biosecurity Flagship}, % university, etc
  \street{11 Julius Av},                     %
  \postcode{2113},                                % post or zip code
  \city{Sydney},                              % city
  \cny{Australia}                                    % country
}
\address[id=aff2]{%
  \orgname{School of Biomedical Sciences and Pharmacy, Faculty of Health},
  \postcode{2308},
  \city{Newcastle},
  \cny{Australia}
}
\address[id=aff3]{%
  \orgname{Cancer Epigenetics Program, Cancer Research Division, Kinghorn Cancer Centre, Garvan Institute of Medical Research},
  \street{384 Victoria St},
  \postcode{2010},
  \city{Sydney},
  \cny{Australia}
}
\address[id=aff4]{%
  \orgname{UNSW Medicine, University of New South Wales},
%  \street{384 Victoria St},
  \postcode{2052},
  \city{Sydney},
  \cny{Australia}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To further demonstrate the scalability of \variantSpark{}, we also cluster the 1000 Genomes Project phase 3 data, which contains 3000 individuals from 5 super-populations and as a result has over 80 Million variants. 
The uncompressed size of the phase 3 files is 770GB compared to the 161GB of the phase 1 dataset.
\variantSpark{} successfully completes the clustering in 27 hours (see Table 2) with an ARI of 0.82.

@@

%TODO add value
\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in XX hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 86\% faster than Hadoop (29 min). %7-fold speedup
\variantSpark{} scales to data sizes that are not feasible to process using R or Python due their requirement to load the whole dataset into memory. 
But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 90\% speedup over R (42 min), and 96\% over Python (103 min). %10 fold %25-fold
\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data. 
Utilising MLlib as well as Spark.ML will enable supervised machine learning applications to e.g. identify variants that jointly interact with phenotypes as well as include electronic health record in addition to the genomic feature vector to e.g. capture medical history as well as predispositions for diagnosis and treatment decisions.

@@

\subsection*{R Implementation}
%For our R implementation, we utilize {\sc read.table} to read in a VCF file as a tab-delimited document. As this also captures the first nine columns that don't contain variants, we subsequently remove those columns with {\sc vcf[1:9] <- list()}.
For our R implementation, we utilize {\sc read.table} to read in a VCF file as a tab-delimited document and remove the first nine columns that do not contain variants. 
%In the supplementary material we benchmark a slightly faster (by less than one minute) approach of reading the VCF file in using the `VariantAnnotation' package.
In the supplementary material we benchmark this against an approach reading the VCF file using the `VariantAnnotation' package, which causes inefficiencies for the subsequent data manipulation tasks and is hence overall slower.

As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value. This process is identical for both of our R approaches, and essentially consists
of applying our {\sc Hamming} function to the dataframe with {\sc sapply}.
We then transpose the matrix with {\sc t(vcfMatrix)}, which results in a data-structure where each row represents an individual. We convert the matrix to a {\sc big.matrix} object,
as required by the \kMeans{} algorithm from the `biganalytics' package (\url{https://cran.r-project.org/web/packages/biganalytics/index.html}), and then call {\sc bigkmeans} with
the {\sc big.matrix} object and the required number of clusters as arguments.

>>>>

\author[
   addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
%   corref={aff1},                       % id of corresponding address, if any
%   noteref={n1},                        % id's of article notes, if any
%   email={jane.e.doe@cambridge.co.uk}   % email address
]{\inits{}\fnm{Aidan R.} \snm{O'Brien}}
\author[
   addressref={aff1},
]{\inits{}\fnm{Neil F.W.} \snm{Saunders}}
\author[
   addressref={aff5},
   email={Yi.Guo@csiro.au}
%   email={john.RS.Smith@cambridge.co.uk}
]{\inits{}\fnm{Yi} \snm{Guo}}
\author[
   addressref={aff3,aff4},
%   email={john.RS.Smith@cambridge.co.uk}
]{\inits{}\fnm{Fabian A.} \snm{Buske}}
\author[
   addressref={aff2},
   email={rodney.scott@newcastle.edu.au}
]{\inits{}\fnm{Rodney J.} \snm{Scott}}
\author[
   addressref={aff1},
   corref={aff1},
   email={Denis.Bauer@CSIRO.au}
%   email={john.RS.Smith@cambridge.co.uk}
]{\inits{}\fnm{Denis C.} \snm{Bauer}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%                           % unique id
  \orgname{CSIRO, Health \& Biosecurity Flagship}, % university, etc
  \street{11 Julius Av},                     %
  \postcode{2113},                                % post or zip code
  \city{Sydney},                              % city
  \cny{Australia}                                    % country
}
\address[id=aff2]{%
  \orgname{School of Biomedical Sciences and Pharmacy, Faculty of Health},
  \postcode{2308},
  \city{Newcastle},
  \cny{Australia}
}
\address[id=aff3]{%
  \orgname{Cancer Epigenetics Program, Cancer Research Division, Kinghorn Cancer Centre, Garvan Institute of Medical Research},
  \street{384 Victoria St},
  \postcode{2010},
  \city{Sydney},
  \cny{Australia}
}
\address[id=aff4]{%
  \orgname{UNSW Medicine, University of New South Wales},
%  \street{384 Victoria St},
  \postcode{2052},
  \city{Sydney},
  \cny{Australia}
}
\address[id=aff5]{%
  \orgname{CSIRO, Data61},
  \postcode{2052},
  \city{Sydney},
  \cny{Australia}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To further demonstrate the scalability of \variantSpark{}, we also cluster the 1000 Genomes Project phase 3 data, which contains 3000 individuals from 5 super-populations and as a result has over 80 Million variants. 
The uncompressed size of the phase 3 files is 770GB compared to the 161GB of the phase 1 dataset.
\variantSpark{} successfully completes the clustering in 30 hours (see Table 2) with an ARI of 0.82.

@@

\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in 30 hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 86\% faster than Hadoop (29 min). %7-fold speedup
\variantSpark{} scales to data sizes that are not feasible to process using R or Python due their requirement to load the whole dataset into memory. 
But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 90\% speedup over R (42 min), and 96\% over Python (103 min). %10 fold %25-fold
\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data. 
Utilising MLlib as well as Spark.ML will enable supervised machine learning applications to e.g. identify variants that jointly interact with phenotypes as well as include electronic health record in addition to the genomic feature vector to e.g. capture medical history as well as predispositions for diagnosis and treatment decisions.

@@

\subsection*{R Implementation}
We utilise the {\sc readGT} from the {\sc VariantAnnotation} package for reading in the VCF file and extracting the the genotype matrix.  
In the supplementary material we demonstrate that this approach is approximately one minute faster than using R's built-in {\sc read.table} function, which requires  removing the first nine columns that do not contain genotypes.

As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value. 
This process consists of applying our {\sc Hamming} function to the dataframe with {\sc sapply}.
We then transpose the matrix with {\sc t(vcfMatrix)}, which results in a data-structure where each row represents an individual. We convert the matrix to a {\sc big.matrix} object,
as required by the \kMeans{} algorithm from the `biganalytics' package (\url{https://cran.r-project.org/web/packages/biganalytics/index.html}), and then call {\sc bigkmeans} with
the {\sc big.matrix} object and the required number of clusters as arguments.
